---
title: "D2_Exercise6"
author: "Dan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Write a for loop that iterates over the integers 1 to 7 and prints the number raised to the power of three.

```{r}
for (i in 1:13) {
  print(i^3)
}
```
Write a for loop that iterates over the names of the columns in the builtin data set iris and prints each
column name together with the number of characters in that column name.
Example output: Sepal.Length: 12
Hint: use the functions print(), paste0(), and nchar().
Remember to read the help page of each function to learn more about them.

```{r}
for (i in 1:5) {
 print(colnames(iris)[i])
 print(nchar(colnames(iris)[i]))
}
```
Answer given:

```{r}
for (column_name in colnames(iris)) {
  print(
    paste0(column_name, ":", nchar(column_name)))
}
```


Use the ifelse() function to print the name of colours that are made up of four characters in the vector
my_colours below.

```{r}
my_colours <- c("red","orange","purple","yellow","pink","blue")

ifelse( test = nchar(my_colours) == 4, yes =my_colours, no = "" )

```

