---
title: "D2_Exercise3"
author: "Dan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the airquality data set.

```{r}
data("airquality")
```

Open the help page for this data set.

```{r}
help("airquality")
```

Examine the data set.

```{r}
airquality
```

Display the column names of the airquality data frame.

```{r}
str(airquality)
```
Sort the data frame by increasing value in the Ozone column.
I will do this by indexing the airquality by row, and defining that index using the order function. I'll then print the new df as a sanity check
```{r}
sort_ozone_aq <- airquality[order(airquality$Ozone),]
sort_ozone_aq
```

Sort the data frame by Month in increasing order and Temp in decreasing order. 

First defining an index to use for sorting. Decreasing can be given a logical vector rather than a single argument. Then using that index to sort the df.
```{r}
sort_index_monthtemp <- order(airquality$Month, airquality$Temp, decreasing = c(FALSE, TRUE)) 

sorted_aq <- airquality[sort_index_monthtemp,]
sorted_aq

```

Write the latest sorted data frame to a text file format of your choice.

```{r}
write.table(sorted_aq, file = "sorted_airquality.tsv", sep = "\t")
```

Run code supplied in powerpoint (should make two dataframes):

```{r}
buildings <- data.frame(
  site = c(1, 2, 3), 
  name = c("b1","b2","b3"))

survey_data <- data.frame(
survey = c("A","A","A","B","B","B"),
location = c(1, 2, 3, 2, 3, 1),
efficiency = c(51, 64, 70, 71, 80, 58))
```

What is the shared information in these two data frames? 

Location information 

Use the merge() function to combine the two data frames by the shared information into a new data frame
called buildings_survey.

```{r}
merged <- merge(x = buildings, y = survey_data, by.x = "site", by.y = "location")
```

Compute the mean of each numeric column each month in the airquality data frame using aggregate().
Make sure NA values are removed.

```{r}
aggregate(x = airquality, 
          by = list(onth = airquality$Month),
          FUN = mean, 
          na.rm = TRUE)
```


Compute the mean of the Solar.R column each month. (also have done this for temp)
Make sure the grouping column is called Month in the return value.
Make sure NA values are removed.

```{r}
aggregate(
	x = airquality[, c("Solar.R", "Temp")],
	by = list(Month = airquality$Month),
	FUN = mean,
	na.rm = TRUE
)
```


